Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-10 scagnostics.Rmd
Footnotes in rmd style rather than latex.

A large number of projection pursuit indexes have been developed, to detect departure from multivariate normal, which includes clusters or outliers or separations between known groups (e.g. @f87, @CBC92, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437).  Less work has been done on indexes to find nonlinear dependence between variables. \footnote{\textcolor{red}{The second component in implementing a projection pursuit method is the optimisation algorithm. Several algorithms have been suggested, some tailored to specific applications, see e.g. XXX references?. In this work our focus is on the index functions, and we will only use optimisation algorithms available via the guided tour implementation of tourr.}}

@@

[@Grimm2016] explored the behavior of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], and computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308], round out the collection of metrics. \footnote{It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in \cite{huber85,JS87}.}
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

## New index functions
\label{sec:indexDef}
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developed for the selection of variable pairs that show interesting features in a scatter plot, but we generalist the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimizing over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:\footnote{We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.}

>>>>

A large number of projection pursuit indexes have been developed, to detect departure from multivariate normal, which includes clusters or outliers or separations between known groups (e.g. @f87, @CBC92, @lckl2005, @AHC02, @CEM:CEM2568, @JS87, @5508437).  Less work has been done on indexes to find nonlinear dependence between variables. ^[\textcolor{red}{The second component in implementing a projection pursuit method is the optimisation algorithm. Several algorithms have been suggested, some tailored to specific applications, see e.g. XXX references?. In this work our focus is on the index functions, and we will only use optimisation algorithms available via the guided tour implementation of tourr.}]

@@

[@Grimm2016] explored the behavior of scagnostics for selecting variables, and proposed two more, based on smoothing splines and distance correlation, that have some nicer properties. In addition the availability of another two indices based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], and computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308], round out the collection of metrics. ^[It is interesting to note that entropy, closely related to the notion of mutual information, has previously been considered as a projection pursuit index when searching for interesting one-dimensional projections, see discussion in @huber85 and @JS87.]
These indices are all available in R packages [@rref]: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr].

@@

## New index functions
\label{sec:indexDef}
Focusing now on $d=2$ dimensional projections, we aim to identify projections that reveal interesting structures or correlations between combinations of variables. We therefore draw on index functions that have been developed for the selection of variable pairs that show interesting features in a scatter plot, but we generalist the notion such that instead of variable pairs we consider any $2$ dimensional projection of the multivariate data distribution. It means that we consider the projection axes as input to the index functions, which will thus score the interestingness of the given projection. What we are optimizing over are then the entries of the $p\times2$ dimensional projection matrix, considering that they obey orthonormality conditions.
Concretely here we consider the following types of index functions:^[We note that all descriptions are idealized, while in practice approximations and binning are used to obtain reasonable computing time, we refer to original references and package documentation for details.]
